---
title: "Network Based Data Analysis"
author: "Sara Baldinelli"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    df_print: default
    fig_caption: yes
    highlight: zenburn
    theme: readable
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: ALS Project
---

# Preliminary steps
load libraries
```{r}
library("GEOquery")
library("useful")
library("plotly")
library("ggfortify")
library("randomForest")
library("RColorBrewer")
library("glmnet") 
library("ROCR")
library("genefilter")
library("caret")
library("e1071")
library("rScudo")
library("igraph")
library("KEGGREST")
library("KEGGgraph")
library("AnnotationDbi")
library("org.Hs.eg.db")
library("pathfindR")
library("biomaRt")
library("RCy3")
```

Import data

```{r}
gse <- getGEO("GSE212131", destdir = ".", getGPL = FALSE)
df <- gse[[1]]
ex <- df@assayData[["exprs"]]


group <- gse[["GSE212131_series_matrix.txt.gz"]]@phenoData@data[["title"]]
sample <- gse[["GSE212131_series_matrix.txt.gz"]]@phenoData@data[["geo_accession"]]
sex <- gse[["GSE212131_series_matrix.txt.gz"]]@phenoData@data[["gender:ch1"]]
meta_data <- data.frame(groups=group, samples=sample, gender=sex)
```

# Initial analysis

```{r}
dim(ex)
# Create a boxplot with customizations
boxplot(ex, 
        main = "Boxplot of Initial Data",
        col = "lightblue", # Set box color
        border = "black", # Set border color
        ylim = c(0, max(ex) * 1.1), # Adjust y-axis limits
        ylab = "Values", # Label for y-axis
        xlab = "Categories", # Label for x-axis
        )
```


## creating vector of labels for further analysis
labels -> plane color labels
groups_labels -> labels with group names
complete_labels -> combination of group + name of the sample
and some others


```{r}
#set labels for pca
labels = c()
group_labels = c()
complete_labels = c()
short_vs_long = c()
ALS_positions = c()
ALS_labels = c()

df$title <- gsub("-[0-9]+", "", df$title)

for (x in 1:42) {
  if (df$title[x] == "Short") {
    labels[x] <- "lightblue"
    group_labels[x] <- "Short"
    complete_labels[x] <- paste(colnames(df)[x], "(Short)")
    short_vs_long[x] <- "0"
    ALS_positions <- c(ALS_positions, x)
    ALS_labels <- c(ALS_labels, "Short")
  } else {
    labels[x] <- "orange"
    group_labels[x] <- "Long"
    complete_labels[x] <- paste(colnames(df)[x], "(Long)")
    short_vs_long[x] <- "1"
    ALS_positions <- c(ALS_positions, x)
    ALS_labels <- c(ALS_labels, "Long")
  }
}
```


## Filtering the dataset by inspecting p-values 
keep only probes that have a significant difference 
-> filtered dataset is called *ex2*

```{r}
# performing t-test
# this test is done for feature selection
# I avoid considering features with a p-value > 0.01

p_values_short_long <- c()

# Convert 'group_labels' to a factor with levels "Short" and "Long" and specify "Long" as the reference level
group_labels <- factor(group_labels, levels = c("Short", "Long"))
group_labels <- relevel(group_labels, ref = "Long")

for (i in 1:22011){
  # for each mRNA expression row ex[i, 0]
  short = c()
  long = c()
  
  for (j in 1:42){
    # for each sample in column j ex[i, j]
    if (group_labels[j] == "Long"){
      # control case
      long <- append(long, 2^ex[i, j])
    } else {
      short <- append(short, 2^ex[i, j])
    } 
  }
  
  t_short_long <- t.test(short, long)
  p_values_short_long <- append(p_values_short_long, t_short_long$p.value)
}

threshold <- 0.01
# create empty new matrix
ex2 = matrix(,nrow = 0, ncol = 42)
ex2_unique_probes = matrix(,nrow = 0, ncol = 42)
colnames(ex2) <- colnames(ex)
colnames(ex2_unique_probes) <- colnames(ex)

for (i in 1 : 22011){
  if (p_values_short_long[i] < threshold) {
    # interesting values
    ex2 <- rbind(ex2, ex[i, ])
    ex2_unique_probes <- rbind(ex2_unique_probes, ex[i, ])
    rownames(ex2)[dim(ex2)[1]] <- rownames(ex)[i] 
    rownames(ex2_unique_probes)[dim(ex2_unique_probes)[1]] <- rownames(ex)[i] 
  }
}
```

# PCA

### performing 2d - PCA with first two components 

```{r}
pca <- prcomp(t(ex2))
screeplot(pca)

components <- pca$x
components <- data.frame(components)
components <- cbind(components, group_labels)

fig <- plot_ly(components, x = ~PC1, y = ~PC2, color = group_labels, colors =  c('#636EFA','orange'), type = "scatter", mode = "markers", title="PCA analysis with whole dataset", symbol = gse[["GSE212131_series_matrix.txt.gz"]]@phenoData@data[["gender:ch1"]])

fig
```

### performing 3d - PCA

```{r}
components <- pca$x
components <- data.frame(components)
components$PC2 <- -components$PC2
components$PC3 <- -components$PC3
components = cbind(components, group_labels)

tot_explained_variance_ratio <- summary(pca)[["importance"]]['Proportion of Variance',]
tot_explained_variance_ratio <- 100 * sum(tot_explained_variance_ratio)

tit = '3D pca plot'
fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = group_labels, colors = c('#636EFA','orange')) %>%
  add_markers(size = 12)
fig <- fig %>%
  layout(
    title = tit,
    scene = list(bgcolor = "white")
)

fig
```

```{r}
# Create 3D scatter plot with customizations
fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = group_labels, colors = c('#636EFA','orange')) %>%
  add_markers(size = 20) %>%  # Increase marker size
  layout(
    title = tit,
    scene = list(
      bgcolor = "white",  # Set background color
      xaxis = list(title = "PC1", range = c(min(components$PC1) - 1, max(components$PC1) + 1)),  # Set label and range for x-axis
      yaxis = list(title = "PC2", range = c(min(components$PC2) - 1, max(components$PC2) + 1)),  # Set label and range for y-axis
      zaxis = list(title = "PC3")  # Set label for z-axis
    )
  )
fig
```


# Clustering

### k-means
```{r}
k <- 2
set.seed(2900)
kmeans_results <- kmeans(t(ex2), k)
table(kmeans_results$cluster)


cluster_colors <- c("#636EFA", "orange")

plot(kmeans_results, data = t(ex2)) +
  geom_text(aes(label = group_labels), size = 3, hjust = -0.15, vjust = 0) +
  theme_bw() +
  scale_color_manual(values = cluster_colors)
```

### hierarchical clustering
```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "ave")

k <- 2
groups = cutree(hc_results, k=k)
plot(hc_results, hang <- 1, labels = group_labels, main = "Hierarchical clustering")
rect.hclust(hc_results, k=k, which = NULL, x = NULL, h = NULL, cluster = NULL, border = cluster_colors)
```

```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "complete")

k <- 2
groups = cutree(hc_results, k=k)
# plot(hc_results, hang <-1, labels = groups)
plot(hc_results, hang <-1, labels = group_labels, main = "Complete hierarchical clustering")
rect.hclust(hc_results, k=k, which = NULL, x = NULL, h = NULL, cluster = NULL, border = 2)
```

```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "single")

k <- 2
groups = cutree(hc_results, k=k)
# plot(hc_results, hang <-1, labels = groups)
plot(hc_results, hang <-1, labels = group_labels, main = "Single hierarchical clustering")
rect.hclust(hc_results, k=k, which = NULL, x = NULL, h = NULL, cluster = NULL, border = 2)
```


# Supervised analysis

## Random forest

random forest error
```{r}
set.seed(2000)
rf <- randomForest(x = t(ex2), y = as.factor(group_labels), ntree=1000)
plot(rf, main = "random forest error",
     xlab = "gene rank", ylab="variable importance",) # tells whether rf does a good job in classifying the 2 groups
plot(sort(rf$importance, decreasing = T))
```


```{r}
# ffun <- filterfun(pOverA(0.20, 0.0))
```


ROW-WISE T-TEST
```{r}
# f <- factor(c(rep(0,22), rep(1,20)))
# tt42 <- rowttests(ex,f)
# 
# keepers <- which(tt42$p.value<0.1)
# 
# tex <- t(ex)
# tex <- tex[, keepers]
# 
# dat <- cbind(tex, c(rep(0,22), rep(1,20)))
# colnames(dat)[ncol(dat)] <- "Long"
# 
# n.short <- 22
# n.long <- 20
# 
# train <- sample(1:(n.short), (n.short-5))
# test <- setdiff(1:(n.short), train)
# 
# test <- c(test, test+22)
# train <- c(train, train+22)
```


gene rank for importance of random forest classification

```{r}
imp.temp <- abs(rf$importance)
t <- order(imp.temp, decreasing = TRUE)

plot(rf, imp.temp[t], log="x", cex.main = 1.5, 
     xlab = "gene rank", ylab="variable importance", cex.lab = 1.5,
     phc = 16, main = "all subset results")
```

heatmap
```{r}
gn.imp <- names(imp.temp)[t]
gn.25 <- gn.imp[1:25]
t <- is.element(rownames(ex2), gn.25)
sig.ex <- ex2[t,]

# make heatmap
hmcol <- colorRampPalette(brewer.pal(11, "GnBu"))(500)
colnames(ex2) <- group_labels
csc <- rep(hmcol[50], 42)
csc[group_labels == "Short"] <- "orange"
csc[group_labels == "Long"] <- "#636EFA"
heatmap(ex2, scale="row", col=hmcol, ColSideColors = csc, main = "Heatmap for gene espression", labCol = F, labRow = F)
```

## LDA, RF and LASSO

trial: how much does it change by losing the control samples? -> enough
```{r}
#subsetting the ex2 with only ALS_long and ALS_short
ex3 <- ex2[, ALS_positions]
ex3_unique_probes <- ex2_unique_probes[, ALS_positions]

set.seed(2000)
rf <- randomForest(x = t(ex3), y = as.factor(ALS_labels), ntree=1000)
plot(rf, main = "random forest error on ALS samples")
```

chec= importance for evALSy probe used in random forest training

```{r}
imp.temp <- abs(rf$importance)
t <- order(imp.temp, decreasing = TRUE)

plot(c(1:nrow(ex2)), imp.temp[t], log="x", cex.main = 1.5, 
     xlab = "gene rank", ylab="variable importance", cex.lab = 1.5,
     phc = 16, main = "Short vs long subset results")
```

Accuracy race between RF, LASSO and LDA classification

```{r}
f <- factor(short_vs_long)
tt40 <- rowttests(ex3, f)
keeps <- which(tt40$p.value<0.01)
tex = t(ex3)
dat <- cbind(as.data.frame(tex), f)

colnames(dat)[ncol(dat)] <- "CONDITION"
n.Short <- 22
n.Long <- 20

control <- trainControl(method="cv", number = 10)
metric <- "Accuracy"

fit.lda <- caret::train(CONDITION~., data=dat, method="lda", metric=metric, trControl=control)

fit.rf <- caret::train(CONDITION~., data=dat, method="rf", metric=metric, trControl=control)

fit.lasso <- caret::train(CONDITION~., 
                          data=dat, 
                          method="glmnet", 
                          family="binomial", 
                          metric=metric, 
                          trControl=control, 
                          tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by=0.05)))
                          
fit.ridge <- caret::train(CONDITION~. ,
                          data=dat, 
                          method = 'glmnet', 
                          faimily = "binomial", 
                          metric = metric, 
                          trControl=control,
                          tuneGrid = expand.grid(alpha = 0, lambda =  seq(0, 1, by=0.05))) 


results <- resamples(list(LDA=fit.lda, RF=fit.rf, LASSO=fit.lasso, RIDGE = fit.ridge))
summary(results)
ggplot(results) + labs(y="Accuracy")
```
testing regression lambda parameters

```{r}
test <- sample(1:(n.Short + n.Long), 5)
train <- setdiff(1:(n.Short + n.Long), test)

# plot of the coefficients for different values of lambda
fit = glmnet(dat[train,], short_vs_long[train], alpha = 0.7,  standardize = F, family = "binomial")
plot(fit, xvar = "lambda", label = T)

# this runs inside the cross validation
cfit = cv.glmnet(data.matrix(dat[train,]), alpha = 0.7, short_vs_long[train], standardize = F, family = "binomial")
plot(cfit)

# occasionally you find params with beta != zero
# coef(cfit, s=cfit$lambda.min)

predict(fit, data.matrix(dat[test,]), type = "class", s=cfit$lambda.min)

pred2 <- predict(fit, data.matrix(dat[test,]), type = "response", s = cfit$lambda.min)
plot(performance(prediction(pred2, short_vs_long[test]), "tpr", "fpr"))
auc.tmp <- performance(prediction(pred2, short_vs_long[test]), "auc")
auc <- as.numeric(auc.tmp@y.values)
```

# rScudo

```{r}
n.top.bottom = 25
N_ = 0.4

set.seed(123)
inTrain <- createDataPartition(f, list = F)
trainData <- ex3_unique_probes[, inTrain]
testData <- ex3_unique_probes[, -inTrain]

trainRes <- scudoTrain(trainData, groups = f[inTrain], 
                       nTop = n.top.bottom, nBottom = n.top.bottom, alpha = 0.05)

upSignatures(trainRes)[1:5, 1:5]
consensusUpSignatures(trainRes)[1:5, ]

# training
trainNet <- scudoNetwork(trainRes, N= N_)
scudoPlot(trainNet, vertex.label = NA)

# validation with test 
testRes <- scudoTest(trainRes, testData, f[-inTrain])
                     # ntop = 25, nbottom = 25)
testNet <- scudoNetwork(testRes, N = N_)
scudoPlot(testNet, vertex.label = NA)

# indentify clusters on map
testClust <- igraph::cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)

# perform classification
classRes <- scudoClassify(trainData, testData, N= N_,
                          nTop = n.top.bottom, nBottom = n.top.bottom,
                          trainGroups = f[inTrain], alpha = 0.5)
caret::confusionMatrix(classRes$predicted, f[-inTrain])
```

## rScudo + caret

```{r}
model <- scudoModel(nTop = 25, nBottom = 25, N = N_)
control <- caret::trainControl(method = "cv", number = 5, summaryFunction = caret::multiClassSummary)
cvRes <- caret::train(x=t(trainData), y=f[inTrain], method = model, trControl = control)

testRes <- scudoTest(trainRes, testData, f[-inTrain],
cvRes$bestTune$nTop,
cvRes$bestTune$nBottom5)
testNet <- scudoNetwork(testRes, N = 0.4)
scudoPlot(testNet, vertex.label = NA)

classRes <- scudoClassify(ex3_unique_probes[, inTrain], ex3_unique_probes[,-inTrain], N_,
                          cvRes$bestTune$nTop, cvRes$bestTune$nBottom,
                          f[inTrain], alpha = 0.05)
caret::confusionMatrix(classRes$predicted, f[-inTrain])
```

preparing lists of genes selected for each method that has been used till now

```{r}
# include in this list the first 200 genes that are considered important for the random forest classification

# rf important genes for classification
rf_importance <- varImp(fit.rf, scale=FALSE)
t <- order(rf_importance$importance$Overall, decreasing = TRUE)
rf_importance_list = c()
for (i in 1:200){
  rf_importance_list <- append(rf_importance_list, 
                               strsplit(rownames(rf_importance$importance)[t[i]], "`")[[1]][2])
}

# lda important genes for classification
lda_importance <- varImp(fit.lda, scale=FALSE)
t <- order(lda_importance$importance$X0, decreasing = TRUE)
lda_importance_list <- rownames(lda_importance$importance[t[1:300], ])

# lasso important genes for classification
lasso_importance <- varImp(fit.lasso, scale=FALSE)
t <- order(lasso_importance$importance$Overall, decreasing = TRUE)
# using different notations i hate it
lasso_importance_list <- c()
for (i in 1:200){
  lasso_importance_list <- append(lasso_importance_list,
                                  strsplit(rownames(lasso_importance$importance)[t[i]], "`")[[1]][2])
}

#coef(fit.lasso, s=fit-lasso$lambda.min)

# ridge important genes for classification
ridge_importance <- varImp(fit.ridge, scale=FALSE)
t <- order(ridge_importance$importance$Overall, decreasing = TRUE)
# they use different notations i hate it
ridge_importance_list <- c()
for (i in 1:200){
  ridge_importance_list <- append(ridge_importance_list,
                                  strsplit(rownames(ridge_importance$importance)[t[i]], "`")[[1]][2])
}

# lapply(ridge_importance_list, write, "/Users/Sara/Desktop/UNI/Lauria/PROJECT/ridge_list.txt", append=TRUE, ncolumns=1000)
# lapply(lda_importance_list, write, "/Users/Sara/Desktop/UNI/Lauria/PROJECT/lda_list.txt", append=TRUE, ncolumns=1000)
# lapply(rf_importance_list, write, "/Users/Sara/Desktop/UNI/Lauria/PROJECT/rf_list.txt", append=TRUE, ncolumns=1000)
# lapply(lasso_importance_list, write, "/Users/Sara/Desktop/UNI/Lauria/PROJECT/lasso_list.txt", append=TRUE, ncolumns=1000)
```


# Network analysis


```{r}
# load ensembl data
ensembl_Mart <- useMart("ensembl")
ensembl <- useEnsembl(biomart = "ensembl", 
                      dataset = "hsapiens_gene_ensembl")
# look for conversion table for probes
ensembl_convertion_table <- getBM(attributes = c('affy_huex_1_0_st_v2', 'hgnc_symbol'),
      filters = 'affy_huex_1_0_st_v2', 
      values = rf_importance_list,
      # values = ridge_importance_list,
      # values = lasso_importance_list,
      # values = lda_importance_list,
      mart = ensembl)
# remove empty labels
ensembl_convertion_table <-  ensembl_convertion_table[!(ensembl_convertion_table$hgnc_symbol == ""), ]

# insert p-value column
ensembl_convertion_table["p_value"] <- NA

# for each selected gene, calculate p-value
for (target in rf_importance_list){
# for (target in ridge_importance_list){
# for(target in lasso_importance_list){
# for(target in lda_importance_list){
  Short = c()
  Long = c()
  for (j in ALS_positions){
    if (group_labels[j] == "Short") {
      Short <- append(Short, 2^ex[target, j])
    } else if (group_labels[j] == "Long"){
      Long <- append(Long, 2^ex[target, j])
    }
  }
  t_short_long <- t.test(Short, Long)
  
  # insert p-value in correct row
  for (i in 1:nrow(ensembl_convertion_table)){
    # print(instance)
    if (ensembl_convertion_table[i, ]$affy_huex_1_0_st_v2 == target){
      ensembl_convertion_table[i, ]$p_value = t_short_long$p.value
    }
  }
}

# run_pathfindR
pathfindR_results <- run_pathfindR(ensembl_convertion_table[, c("hgnc_symbol", "p_value")], iterations = 1)
pathfindR_cluster <- cluster_enriched_terms(pathfindR_results)
term_gene_graph(pathfindR_results,
                layout = "stress",
                use_description = F,
                node_size = "num_genes",
                node_colors = c("orange", "lightblue", "purple"))
enrichment_chart(pathfindR_results)
```
```{r include=FALSE}
# visualize_terms(pathfindR_results, hsa_KEGG = F)
```

```{r include=FALSE}
lapply(ensembl_convertion_table$hgnc_symbol, write, "/Users/Sara/Desktop/UNI/Lauria/PROJECT/nchg_names.txt", append=TRUE, ncolumns=1000)

gene_list <- ensembl_convertion_table$hgnc_symbol
```

```{r}
# 200 probes
library(gprofiler2)
# see vignette at https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
gostres <- gost(query = gene_list, organism = "hsapiens", ordered_query = FALSE,
                multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
                measure_underrepresentation = FALSE, evcodes = FALSE,
                user_threshold = 0.05, correction_method = "g_SCS",
                domain_scope = "annotated", custom_bg = NULL,
                numeric_ns = "", sources = NULL, as_short_link = FALSE)

names(gostres)
head(gostres$result)

# visualize results using a Manhattan plot
gostplot(gostres, capped = TRUE, interactive = TRUE)

# when ready, create publication quality (static) plot + table of interesting terms/pathways
p <- gostplot(gostres, capped = TRUE, interactive = FALSE)
publish_gostplot(p,
                 highlight_terms = c("GO:0005681","GO:0071013","KEGG:03040",
                                     "REAC:R-HSA-72163","CORUM:1181","WP:WP411"), 
                 width = NA, height = NA, filename = NULL)
```


```{r}
#PROBES FROM PATHFINDR RESULTS
library(gprofiler2)
# see vignette at https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
gostres <- gost(query = c("SART1", "hsa03040", "HNRNPC", "HNRNPK", "HNRNPM", "SMNDC1", "BLNK", "hsa05340", "hsa04064", "hsa04662", "hsa04928", "hsa04670", "RHOA", "hsa04071", "hsa04660", "hsa05133", "CFL1", "hsa04666"),
organism = "hsapiens", ordered_query = FALSE,
multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
measure_underrepresentation = FALSE, evcodes = FALSE,
user_threshold = 0.05, correction_method = "g_SCS",
domain_scope = "annotated", custom_bg = NULL,
numeric_ns = "", sources = NULL, as_short_link = FALSE)

names(gostres)
head(gostres$result)

# visualize results using a Manhattan plot
gostplot(gostres, capped = TRUE, interactive = TRUE)

# when ready, create publication quality (static) plot + table of interesting terms/pathways
p <- gostplot(gostres, capped = TRUE, interactive = FALSE)
publish_gostplot(p, 
                 highlight_terms = c("GO:0005681","GO:0071013","KEGG:03040", "REAC:R-HSA-72163","CORUM:1181","WP:WP411"),
                 width = NA, height = NA, filename = NULL)

gostres_DF <- gostres$result
```

```{r}
scudoCytoscape(trainNet)
```

## INCLUDE IN REPORT
TABLE WITH ACCURACY OR SENSITIVITY SPECIFICITY VALUES OF ALL THE METHODS USED
RF BEST METHOD: PUT THE IMPORTANCE GRAPH!!!
ALL THE MATERIAL NOT FITTING: SUPPLEMENTARY MATERIAL

COMPARE LIST OF PATHWAYS AND GENES WITH THE ONE OF THE AUTHORS OF THE DATASET
LITEARTURE SEARCH AND SEE IF TERMS THAT WE OBTAIN ARE ALREADY ASSSOCIATED WITH THE DISEASE WE ARE STUDYING

OUTPUT OF NETWORK-BASED ANALYSIS

REPRODUCIBILITY:PROVIDE ALL THE INFOS NEEDED FOR SOMEONE TO REPLICATE THE RESULTS

SPECIFY THE VERSIONS OF EACH PACKAGE USED, EVEN THE R VERSION
SPECIFY THE PARAMETERS USED FOR THE DIFFERENT FUNCTIONS

5-10 PAGES FOR THE REPORT
10-12 SLIDES (ADD A SLIDE ON CONCLUSION AND FUTURE WORK)
10-12 MINS PRESENTATION
